---
title: "CDS 101 – Final Project Report"
author: "Group Name / Members"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
  html_document:
    css: "gmu-cds101.css"
    toc: true
    toc_depth: 3
    number_sections: true
---

> Replace this template text with your own writing, keeping the headings and overall structure aligned with the **CDS 101 project rubric**.

# 1. Problem Definition

- Clearly state the main problem or question you are investigating.  
- Explain why it is interesting or important.  
- Define the objective(s) of your analysis.  
- Mention any key assumptions.

# 2. Data Acquisition & Description

- Name and source of the dataset(s).  
- How you obtained the data (download link, API, etc.).  
- Brief description of the variables / features (with units if relevant).  
- Approximate size (rows, columns).  
- Any ethical, privacy, or bias considerations (if applicable).

```{r setup-data, echo=FALSE, message=FALSE, warning=FALSE}
# DO NOT ALTER THIS CHUNK
# Set knitr options
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  fig.width = 5,
  fig.asp = 0.618,
  out.width = "70%",
  dpi = 120,
  fig.align = "center",
  cache = FALSE
)
# Load required packages
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(infer))
suppressPackageStartupMessages(library(modelr))
suppressPackageStartupMessages(library(broom))
# Load dataset
college <- read_rds("college.rds") %>%
  type_convert(
    na=combine("", "NA", "PrivacySuppressed")
  )
# Set seed
set.seed(98261936)
```

# 3. Data Cleaning & Preprocessing

Describe the steps you took to make the data usable, for example:

- Handling missing values (drop, impute, etc.).  
- Fixing types (dates, numeric vs. categorical).  
- Removing outliers (if done).  
- Creating new features (feature engineering).  
- Any scaling or encoding.

```{r cleaning, eval=FALSE}
college_reduced<-college%>%
select(
UNITID,
RET_FT4,
UGDS_WHITE,
UGDS_BLACK,
UGDS_HISP,
UGDS_ASIAN,
UGDS_AIAN,
UGDS_NHPI,
REGION
)
```

```{r, cleaning, eval=FALSE}
college_reduced<-college_reduced%>%
rename(
retention_rate=RET_FT4,
pct_white=UGDS_WHITE,
pct_black=UGDS_BLACK,
pct_hispanic=UGDS_HISP,
pct_asian=UGDS_ASIAN,
pct_aian=UGDS_AIAN,
pct_nhpi=UGDS_NHPI,
region =REGION
)
```

```{r, cleaning, eval=FALSE}
college_reduced<-college_reduced%>%
mutate(
  pct_nonwhite=pct_black+pct_hispanic+pct_asian+pct_aian+pct_nhpi
)
```

```{r, cleaning, eval=FALSE}
college_reduced <- college_reduced %>%
  mutate(
    region_name = recode(region,
      `0` = "U.S. Service Schools",
      `1` = "New England",
      `2` = "Mid East",
      `3` = "Great Lakes",
      `4` = "Plains",
      `5` = "Southeast",
      `6` = "Southwest",
      `7` = "Rocky Mountains",
      `8` = "Far West",
      .default = NA_character_
    )
  )
```

```{r, cleaning, eval=FALSE}
college_reduced <- college_reduced %>%
  mutate(
    diversity_group = if_else(
      pct_nonwhite > median(pct_nonwhite, na.rm = TRUE),
      "Higher Diversity",
      "Lower Diversity"
    )
  )
```

# 4. Exploratory Data Analysis (EDA)

The exploratory data analysis focuses on the relationship between diversity and retention rate. The main variables examined are retention_rate, pct_nonwhite, and diversity_group. These graphs and summary statistics help show whether higher diversity schools appear to have different retention patterns than lower diversity schools. 

The summary table compares retention rates between higher diversity and lower diversity universities.

```{r eda-summary, message=FALSE, warning=FALSE}
summary_stats <- college_reduced %>%
group_by(diversity_group) %>%
summarize(
count = n(),
mean = mean(retention_rate, na.rm = TRUE),
median = median(retention_rate, na.rm = TRUE),
sd = sd(retention_rate, na.rm = TRUE),
iqr = IQR(retention_rate, na.rm = TRUE),
min = min(retention_rate, na.rm = TRUE),
max = max(retention_rate, na.rm = TRUE)
)
summary_stats
```

The summary statistics show noticeable differences in retention rates between the two diversity groups. Lower diversity schools had a higher average retention rate of about 0.745 compared to about 0.665 for higher diversity schools. The median retention rate was also higher for lower diversity schools. Higher diversity schools also had greater variability, shown by the larger standard deviation and interquartile range.


The first graph compares the distribution of retention rates between higher diversity and lower diversity schools.

```{r eda-plot-example, message=FALSE, warning=FALSE, fig.cap="Example histogram of a numeric variable (EDA rubric)."}
ggplot(college_reduced, aes(x = retention_rate, fill = diversity_group)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 0.05,
    position = "identity",
    alpha = 0.5
  ) +
  labs(
    title = "Distribution of Retention Rates by Diversity Group",
    x = "Retention Rate",
    y = "Density",
    fill = "Diversity Group"
  )
```
This graph shows that both groups have many schools with retention rates between about 0.6 and 0.9. However, the lower diversity group appears more concentrated around higher retention values, while the higher diversity group has a wider spread and includes more schools with lower retention rates.



# 5. Visualization Quality and Storytelling

Use this section to satisfy the **Visualization Quality** rubric criterion:

- Explain why your plot types are appropriate.  
- Comment on labels, legends, colors, and overall readability.  
- Mention any steps you took to make plots accessible and interpretable.

# 6. Modeling Approach

Explain how you framed the problem and which models you chose:

- Type of task (regression, classification, etc.).  
- Baseline model or heuristic, if used.  
- Main model(s) chosen and why they are appropriate.

# 7. Model Implementation & Evaluation

This corresponds to the **Model Implementation & Evaluation** rubric criterion.

Describe:

- Features used.  
- Data splitting strategy (train/test or cross-validation).  
- Metrics used (accuracy, F1, RMSE, etc.).  
- Tables/plots summarizing performance.  
- A short interpretation of each metric/plot.

```{r modeling, eval=FALSE}
# Example structure:
# set.seed(123)
# train_index <- sample(seq_len(nrow(data_clean)), size = 0.8 * nrow(data_clean))
# train <- data_clean[train_index, ]
# test  <- data_clean[-train_index, ]
#
# model <- glm(target ~ ., data = train, family = binomial)
# preds <- predict(model, newdata = test, type = "response")
# # compute metrics...
```

# 8. Conclusions & Recommendations

Summarize the key takeaways:

- Answer your original research question(s) directly.  
- Highlight the most important patterns or relationships you found.  
- Discuss limitations (data size, bias, missing variables, etc.).  
- Suggest possible extensions or future work.

# 9. Code Quality & Reproducibility

Briefly document how someone else can reproduce your results:

- Which R scripts or Rmd files to run.  
- Any required R packages.  

```{r session-info, echo=FALSE}
sessionInfo()
```

# 10. References

List any references you used, such as:

- Dataset documentation.  
- Research papers or articles.  
- Tutorials or blog posts.

# Appendix (Optional)

Include extra plots, diagnostic checks, or model comparisons if needed.
