---
title: "CDS 101 – Final Project Report"
author: "Group Name / Members"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
  html_document:
    css: "gmu-cds101.css"
    toc: true
    toc_depth: 3
    number_sections: true
---

> Replace this template text with your own writing, keeping the headings and overall structure aligned with the **CDS 101 project rubric**.

# 1. Problem Definition

The problem we are investigating is if diversity has an effect on retention rates in American 4 year college.  I think the question is important because it can show students more information about their college and how that correlates to how oithers feel about it; which can help students make more informed decisions about their college choices. 

Our goal is to find out whether or not higher diversity has a positive or negative correlation if any on retention rate.

Our objectives are to:
Analyze 4 year universities 
Compare their diversity
Compare their retention rates
Analyze any correlation if there is any

Our main assumption is that diversity will have a positive effect on retention rate
- Clearly state the main problem or question you are investigating.  
- Explain why it is interesting or important.  
- Define the objective(s) of your analysis.  
- Mention any key assumptions.

# 2. Data Acquisition & Description

The dataset we are using is called college.rds

We obtained it through a 
download link from canvas. The variables we are using are 
UGDS_WHITE- Total share of enrollment of undergraduate degree-seeking students
who are white
UGDS_BLACK- Total share of enrollment of undergraduate degree-seeking 
students who are black UGDS_HISP- Total
share of enrollment of undergraduate degree-seeking students who are Hispanic UGDS_ASIAN- Total share of enrollment of undergraduate degree-seeking students
who are Asian UGDS_AIAN- Total share of enrollment of undergraduate degree-seeking 
students who are American Indian/Alaska Native 
UGDS_NHPI- Total share of enrollment of undergraduate degree-seeking students
who are
Native Hawaiian/Pacific Islander
RET_FT4- The retention rate of 4 year universities in America
The original dataset has 2228 columns, and 5 columns, our dataset has 6614 rows
and 7 columns. 

- Name and source of the dataset(s).  
- How you obtained the data (download link, API, etc.).  
- Brief description of the variables / features (with units if relevant).  
- Approximate size (rows, columns).  
- Any ethical, privacy, or bias considerations (if applicable).

```{r setup-data, echo=FALSE, message=FALSE, warning=FALSE}
# DO NOT ALTER THIS CHUNK
# Set knitr options
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  fig.width = 5,
  fig.asp = 0.618,
  out.width = "70%",
  dpi = 120,
  fig.align = "center",
  cache = FALSE
)
# Load required packages
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(infer))
suppressPackageStartupMessages(library(modelr))
suppressPackageStartupMessages(library(broom))
# Load dataset
college <- read_rds("college.rds") %>%
  type_convert(
    na=combine("", "NA", "PrivacySuppressed")
  )
# Set seed
set.seed(98261936)
```

# 3. Data Cleaning & Preprocessing

Describe the steps you took to make the data usable, for example:
The first thing we did to make the data more usable was narrowing down the data set and making the 'college_reduced' data set.

Next we renamed the variables that we were using to make them easier to understand.

We worked to create new groupings like pct_nonwhite, which made it easier to show if a school has a higher or lower diversity 

Another column was added to name the regions of the schools, and another to sort the types of diversity as higher or lower.

Another problem that we had to work through was working through NA's in the data analysis section, which is what 'is.na'(Filters out rows with NAs) and 'na.rm = TRUE'(ignores NAs when calculating) are for.



- Handling missing values (drop, impute, etc.).  
- Fixing types (dates, numeric vs. categorical).  
- Removing outliers (if done).  
- Creating new features (feature engineering).  
- Any scaling or encoding.

```{r cleaning, eval=FALSE}
college_reduced <- college %>%
select(
UNITID,
RET_FT4,
UGDS_WHITE,
UGDS_BLACK,
UGDS_HISP,
UGDS_ASIAN,
UGDS_AIAN,
UGDS_NHPI,
REGION
)
```

```{r, cleaning, eval=FALSE}
college_reduced<-college_reduced%>%
rename(
retention_rate=RET_FT4,
pct_white=UGDS_WHITE,
pct_black=UGDS_BLACK,
pct_hispanic=UGDS_HISP,
pct_asian=UGDS_ASIAN,
pct_aian=UGDS_AIAN,
pct_nhpi=UGDS_NHPI,
region =REGION
)
```

```{r, cleaning, eval=FALSE}
college_reduced<-college_reduced%>%
mutate(
  pct_nonwhite=pct_black+pct_hispanic+pct_asian+pct_aian+pct_nhpi
)
```

```{r, cleaning, eval=FALSE}
college_reduced <- college_reduced %>%
  mutate(
    region_name = recode(region,
      `0` = "U.S. Service Schools",
      `1` = "New England",
      `2` = "Mid East",
      `3` = "Great Lakes",
      `4` = "Plains",
      `5` = "Southeast",
      `6` = "Southwest",
      `7` = "Rocky Mountains",
      `8` = "Far West",
      .default = NA_character_
    )
  )
```

```{r, cleaning, eval=FALSE}
college_reduced <- college_reduced %>%
  mutate(
    diversity_group = if_else(
      pct_nonwhite > median(pct_nonwhite, na.rm = TRUE),
      "Higher Diversity",
      "Lower Diversity"
    )
  )
```

# 4. Exploratory Data Analysis (EDA)

The exploratory data analysis focuses on the relationship between diversity and retention rate. The main variables examined are retention_rate, pct_nonwhite, and diversity_group. These graphs and summary statistics help show whether higher diversity schools appear to have different retention patterns than lower diversity schools. 

The summary table compares retention rates between higher diversity and lower diversity universities.

```{r eda-summary, message=FALSE, warning=FALSE}
summary_stats <- college_reduced %>%
group_by(diversity_group) %>%
summarize(
count = n(),
mean = mean(retention_rate, na.rm = TRUE),
median = median(retention_rate, na.rm = TRUE),
sd = sd(retention_rate, na.rm = TRUE),
iqr = IQR(retention_rate, na.rm = TRUE),
min = min(retention_rate, na.rm = TRUE),
max = max(retention_rate, na.rm = TRUE)
)
summary_stats
```

The summary statistics show noticeable differences in retention rates between the two diversity groups. Lower diversity schools had a higher average retention rate of about 0.745 compared to about 0.665 for higher diversity schools. The median retention rate was also higher for lower diversity schools. Higher diversity schools also had greater variability, shown by the larger standard deviation and interquartile range.


The first graph compares the distribution of retention rates between higher diversity and lower diversity schools.

```{r eda-plot-example, message=FALSE, warning=FALSE, fig.cap="Example histogram of a numeric variable (EDA rubric)."}
ggplot(college_reduced, aes(x = retention_rate, fill = diversity_group)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 0.05,
    position = "identity",
    alpha = 0.5
  ) +
  labs(
    title = "Distribution of Retention Rates by Diversity Group",
    x = "Retention Rate",
    y = "Density",
    fill = "Diversity Group"
  )
```
This graph shows that both groups have many schools with retention rates between about 0.6 and 0.9. However, the lower diversity group appears more concentrated around higher retention values, while the higher diversity group has a wider spread and includes more schools with lower retention rates.



# 5. Visualization Quality and Storytelling

The histogram is an appropriate graph because it compares the distribution of retention rates between two sources, we used 'fill = diversity_group' to give different colors to differentiate between higher and lower diversity rates. Setting alpha to 0.5 makes the overlapping bars transparent so you can see both sets at the same time.

The faceted scatterplot graph is the correct graph for looking at the amount of diversity across different regions. We faceted by region name which makes each panel reasonable, and to make the data easier to read we used 'geom_smooth(method = "lm")' to add a trend line to each graph, which shows coorelation.
```{r}
ggplot(college_reduced, aes(x = pct_nonwhite, y = retention_rate)) +
  geom_point(alpha = 0.4, size = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~region_name) +
  labs(
    title = "Diversity vs Retention by Region",
    x = "Diversity (Non-White %)",
    y = "Retention Rate"
  )
```
This code chunk is here to find the mean retention rates of schools with lower and higher diversity, and it shows that lower diversity schools have a higher mean retention rate. 
```{r}
college_reduced %>%
  group_by(diversity_group) %>%
  summarize(mean_retention = mean(retention_rate, na.rm = TRUE))
```
This code chunk is here to fix errors with NA's in the null distribution code chunk, which was causing it to see 3 columns instead of two, and removes the third column that was made up of NA's.
```{r}
college_reduced <- college_reduced %>%
  filter(!is.na(diversity_group))
```
This code chunk exists to create the null distribution, which creates 10,000 differences in means that assume that diversity has no effect on retention rate.
```{r}
null_distribution <- college_reduced %>%
  specify(retention_rate ~ diversity_group) %>%
  hypothesize(null = "independence") %>%
  generate(10000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("Higher Diversity", "Lower Diversity"))
```
this is just here to show the null distribution simulations
```{r}
null_distribution
``` 
This code chunk is here to calculate the actual difference in means(−0.080404), and then store it as 'observed_stat', so we can use it for the p-value code and coorelating graph

```{r}
  observed_stat <- 0.6650291 - 0.7454331
  observed_stat
```
The null distribution plot is a histogram type of graph, and shows what the differences in mean would look like if it was due to chance alone. This graph was to visualize the null distribution, and helps clarify why we got a p-value of one.

```{r}
visualize(null_distribution) +
  shade_p_value(obs_stat = observed_stat, direction = "greater")
```
Since we got a P-value of one we fail to reject the null hypothesis of diversity having no effect on college retention rate.
# 6. Modeling Approach

Explain how you framed the problem and which models you chose:

- Type of task (regression, classification, etc.).  
- Baseline model or heuristic, if used.  
- Main model(s) chosen and why they are appropriate.

# 7. Model Implementation & Evaluation

This corresponds to the **Model Implementation & Evaluation** rubric criterion.

Describe:

- Features used.  
- Data splitting strategy (train/test or cross-validation).  
- Metrics used (accuracy, F1, RMSE, etc.).  
- Tables/plots summarizing performance.  
- A short interpretation of each metric/plot.

```{r modeling, eval=FALSE}
# Example structure:
# set.seed(123)
# train_index <- sample(seq_len(nrow(data_clean)), size = 0.8 * nrow(data_clean))
# train <- data_clean[train_index, ]
# test  <- data_clean[-train_index, ]
#
# model <- glm(target ~ ., data = train, family = binomial)
# preds <- predict(model, newdata = test, type = "response")
# # compute metrics...
```

# 8. Conclusions & Recommendations

Based on the testing of our hypothesis we found no evidence that suggests that diversity has a positive effect on retention rate. Our observed statistic which was -0.080404 showed that schools with higher diversity actually had lower retention rates. This pattern is evident when looking at the facted scatterplots you can see the negative relationship between diversity and retention rates. 

One of our biggest limitations was the amount of NA's in the dataset which could have skewed the dataset, another limitation is that retention rate has more influences than just diversity rate, which means those schools could have external factors effecting retention rate.

In the future this could be done by looking at each racial group separately instead of combining them, and having more control for other variables of external factors, and statistics over a timeline could give more accurate data.



Summarize the key takeaways:

- Answer your original research question(s) directly.  
- Highlight the most important patterns or relationships you found.  
- Discuss limitations (data size, bias, missing variables, etc.).  
- Suggest possible extensions or future work.

# 9. Code Quality & Reproducibility

Briefly document how someone else can reproduce your results:

- Which R scripts or Rmd files to run.  
- Any required R packages.  

```{r session-info, echo=FALSE}
sessionInfo()
```

# 10. References

List any references you used, such as:

- Dataset documentation.  
- Research papers or articles.  
- Tutorials or blog posts.

# Appendix (Optional)

Include extra plots, diagnostic checks, or model comparisons if needed.
