---
title: "CDS 101 – Final Project Report"
author: "Group Name / Members"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    css: "gmu-cds101.css"
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
---

> Replace this template text with your own writing, keeping the headings and overall structure aligned with the **CDS 101 project rubric**.

# 1. Problem Definition

The problem we are investigating is if diversity has an effect on retention rates in American 4 year college.  I think the question is important because it can show students more information about their college and how that correlates to how oithers feel about it; which can help students make more informed decisions about their college choices. 

Our goal is to find out whether or not higher diversity has a positive or negative correlation if any on retention rate.

Our objectives are to:
Analyze 4 year universities 
Compare their diversity
Compare their retention rates
Analyze any correlation if there is any

Our main assumption is that diversity will have a positive effect on retention rate
- Clearly state the main problem or question you are investigating.  
- Explain why it is interesting or important.  
- Define the objective(s) of your analysis.  
- Mention any key assumptions.

# 2. Data Acquisition & Description

The dataset we are using is called college.rds

We obtained it through a 
download link from canvas. The variables we are using are 
UGDS_WHITE- Total share of enrollment of undergraduate degree-seeking students
who are white
UGDS_BLACK- Total share of enrollment of undergraduate degree-seeking 
students who are black UGDS_HISP- Total
share of enrollment of undergraduate degree-seeking students who are Hispanic UGDS_ASIAN- Total share of enrollment of undergraduate degree-seeking students
who are Asian UGDS_AIAN- Total share of enrollment of undergraduate degree-seeking 
students who are American Indian/Alaska Native 
UGDS_NHPI- Total share of enrollment of undergraduate degree-seeking students
who are
Native Hawaiian/Pacific Islander
RET_FT4- The retention rate of 4 year universities in America
The original dataset has 2228 columns, and 5 columns, our dataset has 6614 rows
and 7 columns. 

- Name and source of the dataset(s).  
- How you obtained the data (download link, API, etc.).  
- Brief description of the variables / features (with units if relevant).  
- Approximate size (rows, columns).  
- Any ethical, privacy, or bias considerations (if applicable).

```{r setup-data, echo=FALSE, message=FALSE, warning=FALSE}
# Example: load a dataset (replace with your own)
# library(readr)
# data <- read_csv("data/your_data.csv")
```

# 3. Data Cleaning & Preprocessing

Describe the steps you took to make the data usable, for example:

- Handling missing values (drop, impute, etc.).  
- Fixing types (dates, numeric vs. categorical).  
- Removing outliers (if done).  
- Creating new features (feature engineering).  
- Any scaling or encoding.

```{r cleaning, eval=FALSE}
# Example pseudo-code for cleaning:
# data_clean <- data |>
#   dplyr::filter(!is.na(target_variable)) |>
#   dplyr::mutate(
#     new_feature = some_transformation(old_feature)
#   )
```

# 4. Exploratory Data Analysis (EDA)

This section maps directly to the **EDA** part of the rubric:

- Summary statistics  
- Visualizations  
- Interpretation connected to the research question  

```{r eda-summary, message=FALSE, warning=FALSE}
# library(dplyr)
# summary(data_clean)
```

```{r eda-plot-example, message=FALSE, warning=FALSE, fig.cap="Example histogram of a numeric variable (EDA rubric)."}
# library(ggplot2)
# ggplot(data_clean, aes(x = some_numeric_variable)) +
#   geom_histogram(binwidth = 5, color = "white") +
#   labs(x = "Some Variable", y = "Count", title = "Distribution of Some Variable")
```

# 5. Visualization Quality and Storytelling

Use this section to satisfy the **Visualization Quality** rubric criterion:

- Explain why your plot types are appropriate.  
- Comment on labels, legends, colors, and overall readability.  
- Mention any steps you took to make plots accessible and interpretable.

# 6. Modeling Approach

Explain how you framed the problem and which models you chose:

- Type of task (regression, classification, etc.).  
- Baseline model or heuristic, if used.  
- Main model(s) chosen and why they are appropriate.

# 7. Model Implementation & Evaluation

This corresponds to the **Model Implementation & Evaluation** rubric criterion.

Describe:

- Features used.  
- Data splitting strategy (train/test or cross-validation).  
- Metrics used (accuracy, F1, RMSE, etc.).  
- Tables/plots summarizing performance.  
- A short interpretation of each metric/plot.

```{r modeling, eval=FALSE}
# Example structure:
# set.seed(123)
# train_index <- sample(seq_len(nrow(data_clean)), size = 0.8 * nrow(data_clean))
# train <- data_clean[train_index, ]
# test  <- data_clean[-train_index, ]
#
# model <- glm(target ~ ., data = train, family = binomial)
# preds <- predict(model, newdata = test, type = "response")
# # compute metrics...
```

# 8. Conclusions & Recommendations

Summarize the key takeaways:

- Answer your original research question(s) directly.  
- Highlight the most important patterns or relationships you found.  
- Discuss limitations (data size, bias, missing variables, etc.).  
- Suggest possible extensions or future work.

# 9. Code Quality & Reproducibility

Briefly document how someone else can reproduce your results:

- Which R scripts or Rmd files to run.  
- Any required R packages.  

```{r session-info, echo=FALSE}
sessionInfo()
```

# 10. References

List any references you used, such as:

- Dataset documentation.  
- Research papers or articles.  
- Tutorials or blog posts.

# Appendix (Optional)

Include extra plots, diagnostic checks, or model comparisons if needed.
